---
title: "Capstone Project: Analysis of Media Coverage and Sentiment Around Endometriosis with the GDELT API"
author: "Salome Schwegler"
format: pdf
editor: visual
---

## LLM disclosure

## Introduction

Endometriosis is a chronic inflammatory illness in which tissue resembling the endometrial lining of the uterus grows elsewhere, causing severe pain and a wide range of symptoms (Culley et al. 2013). Despite affecting an estimated 190 million people worldwide, endometriosis remains significantly underdiagnosed, with patients waiting an average of seven to ten years before receiving an accurate diagnosis (Shafrir et al. 2018). The estimated number of unreported cases is assumed to be much greater. The diagnostic delay can be attributed to structural shortcomings within healthcare systems and research, as well as to a broader lack of cultural acceptance of the illness (Armour et al. 2020). Both structural issues and the degree of societal acceptance are sustained and reflected by media coverage.

This capstone project, conducted within the data mining class, examines global and multilingual news coverage of endometriosis with the use of the Global Database of Events, Language, and Tone (GDELT Project). The GDELT Project is the largest open database of the human society that has ever been created. The database aims to display all data about our society and to foster interconnectivity across the globe. This is achieved by capturing and representing media coverage from almost every corner of the world. The database is updated daily, while the real-time datasets are updated every 15 minutes.

Analysing news coverage of endometriosis contributes to a better understanding of how the condition is perceived in society and the extent to which new scientific developments, such as advances in identifying the root causes of the illness or emerging treatment options and new medications, are reflected in public discourse.

Truthfully this project aimed to examine the content of news articles and the specific topics covered in articles about endometriosis. This included, for instance, whether articles provide general explanations of the condition, document personal live experiences of patients, report on new medications and treatments in light of recent scientific and academic progress or serve as practical guides for those affected. Just to name a few expected topics. However, this goal proved unfeasible give the scope of the project, the limitations of the API and legal constraints. The GDELT API does not permit the extraction of full article text, partly because the articles originate from a wide range of news organisations and portals, the majority of which operate behind paywalls. It was therefore only possible to retrieve basic metadata via the API, such as the URL, date, country and language of each article. While web scraping could theoretically have enabled full text extraction, given that article URLSs were correctly extracted and hence available, this approach was however deliberately avoided. As discussed repeatedly in class, web scraping requires extreme caution and should be avoided where possible due to legal concerns. It should also be noted that the GDELT API only allows articles extraction for the preceding three months, which poses on additional limitation on the scope of this research. The research question were therefore revised iteratively to reflect what is empirically feasible within the constraints of the data source.

### Research question

Due to the limited scope of this project and the constraints of data accessibility through the GDELT Project API, two adapted research questions will guide the following analysis:  

**RQ1:** Which countries produce the greatest volume of news coverage about endometriosis?

**RQ2:** How has the average sentiment tone of endometriosis news coverage changed over the past four years?

## Data and Methodology

### Data source

All data for this project was retrieved from the GDELT Project. Its document version 2.0. GDELT is a real-time open data platform that monitors print, broadcast and web news media across the world in 152 languages. 65 languages are translated in real-time. It processes hundreds of thousands of articles per day and encodes information about “people, locations, organisations, themes, sources, emotions, counts, quotes, images and events” (<https://www.gdeltproject.org/>). For the purpose of this project two distinct query modes of the GDELT DOC API were used:

“Artlist”, which returns metadata of individual article and “TimelineTone”, which returns time-series sentiment data aggregated across all matching documents.

The API was accessed via HTTP GET request using the httr package in R. The endpoint used was <https://api.gdeltproject.org/api/v2/doc/doc>, with query parameters for the search term, mode, time range and output format. No API key is required to access GDELT, which is a big advantage. However the platform enforces rate limits on the frequency of the requests which is why Sys.sleep(6) and timeout(60) was integrated into the request.

The next section will elaborate on how data was accessed, preprocessed and analysed for both distinct query modes separately.

### Data access and API

The GDELT Project provides open access to its data through an API. An API is a standardised communication channel that allows to request data from one software system to another over the internet. In practical terms, a request is sent to the GDELT server in the form of a URL containing a set of parameters which specify i.e. the research keyword(s), the time range, query mode and the output format. The server then processes the request and returns the matching data which in this case is in JSON format. The JSON format can be parsed and converted into a dataframe in R. All requests were made using the httr package in R, which handles the HTTP GET requests. Additionally as a precaution and for courtesy reasons, Sys.sleep(6) pauses R for 6 seconds before sending the request to avoid being blocked by the server and Timeout(60) which returns an error message if the server has not responded within 60 seconds, are integrated into the request.

In a next step content(response, "text", encoding = "UTF-8") was used to extract the raw response body from the response object. In order for R to read the data properly the fromJSON() function is used to parse the JSON string and convert it into an R list. The articles are now accessible under data\$articles and with the as.data.frame() function converted into a dataframe in R. The dataframe is saved to the data_raw folder for further processing and for practicability so that data does not need to be re-fetched from the API every time the script runs.

### Load libraries

The following R libraries need to be loaded (or installed):

```{r}
# load libraries ----------------------------------------------------------
library(tidyverse)
library(httr)
library(rvest)
library(jsonlite)
library(lubridate)
```

### Article extraction (artlist mode)

To address RQ1, individual article metadata was retrieved using the “artlist” mode. A single API call was first constructed to verify the structure of the response and whether it was possible to get data and to get familiar with the data. A critical limitation of the artlist mode is that it caps results at 250 articles per request. The code only worked when loading the whole code from Sys.sleep(6) until where the data is parsed at the same time and only once per session. Otherwise I was blocked from the server.

#### Getting the data

```{r}
# get 250 articles about endometriosis from the past 3 months
Sys.sleep(6)
response <- httr::GET(
  url = "https://api.gdeltproject.org/api/v2/doc/doc",
  query = list(
    query = "endometriosis",
    mode = "artlist",
    maxrecords = 250,
    timespan = "1week",
    format = "json"
  ),
  timeout(60)
)


# convert data from json to R list
data <- fromJSON(content(response, "text", encoding = "UTF-8"))

#Convert to a clean data frame
articles_df <- as.data.frame(data$articles)

# save articles
#saveRDS(articles_df, "data_raw/articles_df.rds")

```

In order to work around the maximum record limit of 250 articles per request, a day-by-day loop was written to retrieve up to 250 articles for each day across a period of three months (between 25.12.2025 and 24.03.2026). By sending one request per day, the 250-record cap applies independently to each day rather than to the entire period, maximising the total number of articles retrieved. Each request is wrapped in a tryCatch() function, which catches any timeout or connection issues and skips to the next day rather than stopping the loop entirely if a single request fails. The tryCatch() function consists of two blocks. The first block executes the GET request, with two timeout parameters: timeout(60), which controls how long R waits to receive the full response and config(connecttimeout = 60), which controls how long R waits to establish the initial connection to the server. If the request fails, the execution jumps to the second block and prints “Timeout on” the current date and skips that specific day. If a valid response in the first block is received, which is indicated by the HTTP 200 status code, the JSON is parsed and converted into a dataframe. A six-second pause at the end of each iteration ensures the request frequency stays within the GDELT’s limits. Once the loop has iterated through each day, all daily dataframes are combined into a single table using bind_rows().

```{r}
# Define dates
date_sequence <- seq(as.Date("2025-12-25"), as.Date("2026-03-24"), by="day")
all_articles <- list()

for(i in 1:length(date_sequence)){
  current_day <- format(date_sequence[i], "%Y%m%d")
  print(paste("Fetching data for:", current_day))
  
  # tryCatch prevents the code from stopping if there is a timeout error
  result <- tryCatch({
    GET(
      url = "https://api.gdeltproject.org/api/v2/doc/doc",
      query = list(
        query = "endometriosis",
        mode = "artlist",
        maxrecords = 250,
        startdatetime = paste0(current_day, "000000"),
        enddatetime = paste0(current_day, "235959"),
        format = "json"
      ),
      # These two lines fix the timeout, section written with help of AI
      timeout(60),                      # guards the download phase
      config(connecttimeout = 60)       # guards the connection phase
    )
  }, error = function(e) {
    message(paste("Timeout on", current_day, "- skipping."))
    return(NULL)
  })
  
  # Process only if result is valid
  if (!is.null(result) && status_code(result) == 200) {
    raw_content <- content(result, as = "text", encoding = "UTF-8")
    if (!grepl("Please limit requests", raw_content)) {
      data <- fromJSON(raw_content)
      if (!is.null(data$articles)) {
        all_articles[[i]] <- as.data.frame(data$articles)
      }
    }
  }
  
  # Stay under the rate limit
  Sys.sleep(6)
}

# Combine all days into one table
final_endo_df_2 <- bind_rows(all_articles)
final_endo_df_2

# save variable data set
#saveRDS(final_endo_df_2, "data_raw/final_endo_df_2.rds")
```

#### Preprocessing the data

The following code cleans the raw data file and stores the dataframe in the preprocessed data folder for future analysis. The raw date column “seendate” is stored in a compact ISO 8601 format (i.e. 20260326T094500Z). The object has to be parsed into a proper datetime object in R. This is done with the ymd_hms() function from the “lubridate” package. The date is then divided into individual time components such as year, month, day and hour. The individual components are then stored in separate columns, making it straightforward to group and aggregate the data by any of these units in future analysis. The preprocessed data is then saved in the preprocessed folder.

```{r}
# preprocess data for media coverage analysis by country ------------------
# load data
endo_df_preprocessed <- readRDS("data_raw/final_endo_df.rds")

# create readable format of date with  (this step was done with the help of AI)
endo_df_preprocessed <- endo_df_preprocessed |> 
  mutate(date_parsed = ymd_hms(seendate, tz = "UTC"))

# split up date for analysis of years and months and days
endo_df_preprocessed <- endo_df_preprocessed |>
  mutate(
    year  = year(date_parsed),
    month = month(date_parsed, label = TRUE),
    day   = day(date_parsed),
    hour = hour(date_parsed)
  )

# save preprocessed df
saveRDS(endo_df_preprocessed, "data_preprocessed/endo_df_preprocessed.rds")
```

#### Analysis

### Sentiment score (TimelineTone mode)

#### Getting the data

#### Preprocessing the data

#### Analysis

## Conclusion and Limitations
